---
output: github_document

  
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CarefullyCausal

<!-- badges: start -->
<!-- badges: end -->


The goal of CarefullyCausal is to provide the user a practical guide when doing causal analyses. Particularly, CarefullyCausal provides the user the estimand, a table of causal estimates, a discussion on the causal assumptions, relevant diagnostics and interpretations/explanations. The key aspects of a causal analyses are printed and discussed in detail to help the user in evaluating whether the estimated effects can be interpreted as being causal. Currently, CarefullyCausal can be used in a setting with a fixed-exposure, meaning that the exposure does not vary over time.\

Some key features:\

* **Setting**: Fixed-exposure
* **Outcome of interest**: can be dichotomous or continuous
* **Exposure of interest**: can be dichotomous, multi-value (max 4 levels) or continuous
* **Effect measures**: can be in log(odds), risk ratio or odds ratio



## Installation

You can install the development version of CarefullyCausal as follows:
```{r install, eval=FALSE}
# To download R packages from Github or other sources we need the "devtools" package
install.packages("devtools")
library("devtools")

# Now we can download the CarefullyCausal package
install_github("mauricekorf/CarefullyCausal")
library(CarefullyCausal)
```


## Example

To develop some intuition with the CarefullyCausal function and to highlight some important features, an example is step-by-step illustrated using the NHEFS^[National Health and Nutrition Examination Survey Data I Epidemiologic Follow-up Study, https://wwwn.cdc.gov/nchs/nhanes/nhefs/] data set.


### Research Question

Suppose that we are interested in the causal relation between quitting smoking and weight change (in Kilograms). Specifically,we would like to know the effect of quitting smoking on someone's weight change. In order to evaluate this causal relation, we assume that there are some important confounders we need to adjust for, which includes:\
sex, race, age, education level, smoke intensity (number of cigarettes per day), how long someone has been smoking ( in years), start weight (in Kilograms), how much someone exercises (**0:** much exercise, **1:** moderate exercise, **2:** no exercise) and how active a person is on a usual day (**0:** very active, **1:** moderately active, **2:** inactive)

<br>
Given that we assume that these are the only confounders and that no collider bias or selection bias is induced, we would obtain the following simple Directed Acyclic Graph (DAG):
<p align='center'>
```{r RQ, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=10, out.width='60%', fig.align='center'}
library(ggdag)
library(ggplot2)
library(causaldata)
library(kableExtra)
library(CarefullyCausal)

# Load in the data, we will use the complete cases variant
df = nhefs_complete

# Make the coordination of the different nodes
# https://lfoswald.github.io/2021-spring-stats2/materials/session-3/03-online-tutorial/
coord_dag <- list(
  x = c(qsmk = 0, wt82_71 = 3, sex = 1.5, race = 1.5, education = 1.5, smokeintensity = 1.5, smokeyrs = 1.5, wt71 = 1.5, exercise = 1.5, active = 1.5, age = 1.5 ),
  y = c(qsmk = 0, wt82_71 = 0, sex = 1, race = 1.5, education = 2, smokeintensity = 2.5, smokeyrs = -1, wt71 = -1.5, exercise = -2, active = -2.5, age = -3)
)
# Create the DAG structure
our_dag <- ggdag::dagify(wt82_71 ~ qsmk,
                         wt82_71 ~ sex + race + education + smokeintensity + smokeyrs + wt71 + exercise + active + age,
                         qsmk ~ sex + race + education + smokeintensity + smokeyrs + wt71 +exercise + active + age,
                         coords = coord_dag,
                         labels = c("wt82_71" = "Weight\n change",
                                    "qsmk" = "Quit smoke",
                                    "race" = "race",
                                    "sex" = "sex",
                                    "education" = "education",
                                    "smokeintensity" = "number cigarettes",
                                    "smokeyrs" = "years smoking",
                                    "wt71" = "start weight",
                                    "exercise" = "exercise",
                                    "active" = "active",
                                    "age" = "age"))
# Plot the DAG
ggdag(our_dag, text = FALSE, use_labels = "label") + theme_void()

```
</p>
<center>
<i>Figure 1: </i>The selected variables are for illustration purposes only. The DAG shows by no means the true causal structure
</center>
<br>



### Data

In order to answer our research question, we will use a subset of the NHEFS data set from Hernán MA & Robins JM (2020) in the *Causal inference: What if* book^[Hernán MA, Robins JM (2020). Causal Inference: What If. Boca Raton: Chapman & Hall/CRC,https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/]. To conveniently import this data set we will use the package *causaldata*.

```{r download data, eval=FALSE}
# Download the required package, containing various (causal) data sets
install.packages("causaldata")
library(causaldata)

# Load in the data, we will use the complete cases variant
df = nhefs_complete
```
<br>


We will now select the variables that we deemed to be relevant, as shown in Figure 1 (DAG). We show the first six rows of the data set below, to get an idea of the data set.
```{r select variables data, eval=TRUE}
# Select variables
df = nhefs_complete[,c("wt82_71","qsmk","race","sex","education","smokeintensity", "smokeyrs","wt71","exercise","active", "age")]
```

```{r show variables data, eval=TRUE, echo=FALSE}
# Get an idea of the data (first 6 rows)
knitr::kable(head(df))
```
<br>

To further inspect the variable coding and the corresponding definition we will create a table including the class of the variable.
<center>
```{r explain variables data, eval=TRUE, echo=FALSE, fig.align='center', out.width="60%"}
# Get an idea how the variables are coded and what type it is
variables <- setNames(stack(sapply(df, class))[2:1], c('variable', 'class'))
description <- c("The weight change between 1971 and 1982 in Kg ",
                 "Quit smoking, 1: yes, 0: no",
                 "1: black or other, 0: white, ",
                 "1: female, 0: male",
                 "1: 8th grade, 2: HS dropout, 3: HS, 4: college dropout, 5: college or higher",
                 "Number of cigarettes smoked per day in 1971",
                 "Years of smoking",
                 "Start weight (Kg) in 1971",
                 "In recreation in 1971 how mu h exercising, 0: much, 1: moderate, 2: little or none",
                 "On usual day how active in 1971, 0: very active, 1: moderately active, 2: inactive",
                 "Age in 1971")

variables = cbind(variables,description)
knitr::kable(variables)
```
</center>
<br>


### Analysis (Applying CarefullyCausal)

We will now shift our focus to actually using the CarefullyCausal function. The minimal call requires us to specify the following arguments: `formula, data, family and exposure`. Note that you can always consult the help file within R to see the documentation of CarefullyCausal and to learn about all available arguments including an explanation. You can access this by simply typing in *CarefullyCausal* in the help tab in R. Nonetheless, we summarise the key arguments we need for the minimal call.


* `Formula`, this has the same form as for example when using *glm* which is $y\sim x + w$. Here $y$ denotes the outcome of interest variable, $x$ denotes the exposure and $w$ is a covariate we want to adjust for
* `Data`, specify the data frame (data set)
* `Family`, this is the same as in *glm* where it describes the error distribution and link function. Here we have two choices: `"gaussian"` or `"binomial"`. This depends on the nature of the outcome of interest. In our example the outcome is continuous and thus we use the `"gaussian"` argument (default setting)
* `Exposure`, we explicitly define our exposure variable as a character string. In this example, it will be `"qsmk"`


```{r analysis, eval=TRUE, fig.show='hide', warning=FALSE, message=FALSE}
# Transform exposure in a factor variable and transform into dataframe
df$qsmk = as.factor(df$qsmk) 
df = as.data.frame(df)

# Run the CarefullyCausal function and save it in the object "output"
output <- CarefullyCausal(wt82_71 ~ qsmk + race + sex + education + smokeintensity + smokeyrs + wt71 + exercise
                          + active + age,
                          data = df,
                          exposure = "qsmk",
                          family = "gaussian")
# Print the output
output


```
<br>

The output of the CarefullyCausal function is shown above and can be broadly classified into the sections: estimand,
estimators and assumptions/diagnostics. We go over each part now in more detail.

##### Estimand
The first lines in the output show the *estimand*, which is a precise description of your research question. Ideally, this should be decided on before the analysis. Nonetheless, by showing the estimand it may ideally help you double-check your research question and whether this is in line with what you are estimating as shown by the estimand. Moreover, it should be noted that the estimand is in counterfactual notation and thus explicitly emphasizes that we are interested in causal effects. We provide both a conditional and marginal estimand since we provide different estimators and different type of estimands correspond to different estimators. In essence, the idea of the estimand is the same but due to more technical reasons (how the estimators work) we write the estimand in a conditional or marginal way. Important to note is that the marginal estimand is still adjusting for all the variables like shown in the conditional estimand, but we do not write the adjusted variables explicitly in the marginal estimand. To learn more about estimands and its formulation: ^[Luijken, K., van Eekelen, R., Gardarsdottir, H., Groenwold, R. H., & van Geloven, N. (2023). Tell me what you want, what you really really want: Estimands in observational pharmacoepidemiologic comparative effectiveness and safety studies. Pharmacoepidemiology and Drug Safety.] ^[Goetghebeur, E., le Cessie, S., De Stavola, B., Moodie, E. E., Waernbaum, I., & ” the topic group Causal Inference (TG7) of the STRATOS initiative. (2020). Formulating causal questions and principled statistical answers. Statistics in medicine, 39(30), 4922-4948.] ^[Kahan, B. C., Cro, S., Li, F., & Harhay, M. O. (2023). Eliminating ambiguous treatment effects using estimands. American Journal of Epidemiology, 192(6), 987-994.]
```{r estimand output, eval=FALSE}
#> Estimand: 
#> Conditional 
#> E[wt82_71^qsmk=1|race, sex, education, smokeintensity, smokeyrs, wt71, exercise, active, age]  - E[wt82_71^qsmk=0|race, sex, education, smokeintensity, smokeyrs, wt71, exercise, active, age]
#> 
#> Marginal 
#> E[wt82_71^qsmk=1]  -  E[wt82_71^qsmk=0]
#> *Please see output at $Estimand_interpretation for details 

```
<br>

We can obtain a description of the estimand from the saved output, in this example the CarefullyCausal output was saved in the object named *output*. Hence, we call the specific output object.

```{r estimand explanation output, eval=FALSE}
# Show saved output from $Estimand_interpretation
output$Estimand_interpretation

#> [1] "The estimand shows the average causal effect in the population of interest given the different exposure
#> regimes. More specifically, the effect of receiving exposure level 1 with respect to the reference exposure
#> level 0. When adjusting for a set of covariates (race, sex, education, smokeintensity, smokeyrs, wt71,
#> exercise, active, age) the estimand is displayed in either a conditional or marginal way. Particularly,
#> defining the estimand depends on the approach/model used. For example, outcome regression has a conditional
#> estimand while IPTW has a marginal estimand. The T-standardization approach also has a marginal estimand
#> while for the S-standardization it depends on whether interactions between treatment and covariates are
#> taken into account. When no interactions are considered, then it is the same as outcome regression and
#> thus has a conditional estimand, however when all interactions are considered then it is like
#> T-standardization and thus has a marginal estimand."

```
<br>


##### Estimators
After the estimand, the estimates are shown in a table including the standard errors, s-value and 95% confidence interval. The p-value can also be shown by setting `pvalue=TRUE`, but is hidden by default. The S-value is a transformation of the p-value by applying $-log_2(p-value)$. The S-value is continuum and ranges from 0 to infinity, where a p-value of 1 corresponds to a S-value of 0 and a p-value of 0 corresponds to a S-value approaching infinity. Intuitively, the S-value captures the amount of information in the data against the model and quantifies how *suprised* we can be about a specific outome. In other words, a very high S-value implies that we would be suprised by the findings we found, given all background assumptions and test hypotheses. Showing the S-value by default rather than the p-value is motivated by that in practice a lot of studies dichotomize settings based on a single cut-off value (e.g. $0.05$) where below that value it is interpreted as *useful* and above it is interpreted as *not useful*. In this way the focus is generally purely on the statistical significance and tends to result in underacknowledgment of the practical significance.^[Rose-Nussbaumer, J. (2021). Statistical Significance vs Clinical Significance—That Is the Question. JAMA ophthalmology, 139(11), 1235-1235.] ^[Greenland, S., Senn, S. J., Rothman, K. J., Carlin, J. B., Poole, C., Goodman, S. N., & Altman, D. G. (2016). Statistical tests, P values, confidence intervals, and power: a guide to misinterpretations. European journal of epidemiology, 31, 337-350.] ^[14.	Greenland, S. (2019). Valid p-values behave exactly as they should: Some misleading criticisms of p-values and their resolution with s-values. The American Statistician, 73(sup1), 106-114]

CarefullyCausal currently supports five different estimators, as shown below, where key characteristics are listed below:

* *Outcome Regression*: for outcome regression `glm` is implemented. When `family = "gaussian"` linear regression is by default performed and when `family = "binomial"` logistic regression will be implemented.
* *Inverse Probability Treatment Weighting (IPTW)*: Propensity scores are estimated using the `CBPS` package, which are *Covariate Balancing Propensity Scores*^[Fong C, Ratkovic M, Imai K (2022). _CBPS: Covariate Balancing Propensity Score_. R package version0.23, <https://CRAN.R-project.org/package=CBPS>.]. Technical details about how *CBPS* works can be found for binary exposure^[Imai, K., & Ratkovic, M. (2014). Covariate balancing propensity score. Journal of the Royal Statistical Society Series B: Statistical Methodology, 76(1), 243-263.], for multi-value exposure^[Imai, K., & Ratkovic, M. (2015). Robust estimation of inverse probability weights for marginal structural models. Journal of the American Statistical Association, 110(511), 1013-1023.] and for continuous exposure^[Fong, C., Hazlett, C., & Imai, K. (2018). Covariate balancing propensity score for a continuous treatment: Application to the efficacy of political advertisements. The Annals of Applied Statistics, 12(1), 156-177.]. Based on the estimated propensity scores, *non-stabilized* weights (*IP weights*) will be computed. Subsequently, the `svyglm` function from the `survey` package is used to run a weighted glm with the IP weights to get the correct corresponding standard errors. By default `CBPS` is implemented, but the user can also input their own *IP weights* as a vector using the `ip_weights_iptw` argument and offers fully flexibility.
* *S-and-T-standardization*: these two approaches are generally referred within the causal inference literature to the broad approach of *standardization* (G-computation). In CarefullyCausal an explicit distinction is made between different standardization approaches, hence the two different namings. The names are inspired by the machine learning literature where similar ideas are generalized and clear distinctions are made between different approaches such as the S-learner and T-learner^[Künzel, S. R., Sekhon, J. S., Bickel, P. J., & Yu, B. (2019). Metalearners for estimating heterogeneous treatment effects using machine learning. Proceedings of the national academy of sciences, 116(10), 4156-4165.]. The S-learner and T-learner can be viewed as two different standardization approaches where the S-learner uses a single model and the T-learner uses two models. Specifically, with the S-learner a single model is fitted on all data and then the counterfactual effects are predicted under all exposure levels. In contrast, the T-learner first splits the data based on the exposure variable and then fits a model per exposure level. Each model is then used to predict the counterfactual effects for the entire sample under each exposure level. To make a clear link to the causal inference terminology, we introduced the names S-standardization and T-standardization. A user can add interactions in the S-standardization model between a covariate and exposure using the `interaction` argument. When no interactions are included it will be equivalent to outcome regression and when all possible interactions are considered between the adjustment set and exposure it is equivalent to T-standardization. Currently, CarefullyCausal only supports `glm`. Bootstrapping is used to obtain standard errors and the p-value and are controlled by arguments `boot1` and `boot2`. Moreover, a normal confidence interval or bias-corrected accelerated confidence interval can be computed using `confidence`. These estimators do not support a continuous exposure.
* *Targeted Maximum Likelihood Estimation*: 








```{r estimator, eval=FALSE}
#> Treatment effect: 
#>                       Estimate Std. Error S-value 95%.CI.lower 95%.CI.upper
#> qsmk1 outcome regression    3.381      0.441  44.858        2.517        4.246
#> qsmk1 IPTW                  3.318      0.494  35.198        2.351        4.286
#> qsmk1 S-standardization     3.381      0.549     Inf        2.194        4.344
#> qsmk1 T-standardization     3.448      0.523     Inf        2.479        4.528
#> qsmk1 TMLE                  3.370      0.494     Inf        2.401        4.339

#> Reference exposure level: 0 


#> Please evaluate whether the difference beteen the lowest estimate: 3.3183 and highest: 3.4482 is of substance, 
#> given the nature of the data. If so, evaluate the different modelling assumptions.

```
<br>





